### 2015-07-06 narrative/data-reduction.tex
Pushing first cut of editing of data reduction.

\subsection{Data Refactoring}
\label{sec:data-refactor}
It is critical that users give us more information about writing and reading from the SSIO layer
in order to not only gain performance and predictability, but to also guide them to write and read
the best possible data in a given period of time. In our example from C. S. Chang, along with most
of the other scientist we work with, there is a prioritization of data, and this information must be
sent over to our SSIO layer so that the best possible data is delivered to and from the SSIO layer, 
and these decisions often involve refactoring (re-organizing and reducing the data) . In this section we
describe the common methods to reduce data and to re-organize data. Since there are infinite possibilities 
we need to understand what kind of data re-organization will work and when will it work?  We also need to 
understand if it would take too much time and resources to perform the ``best methods'' and how can we manage the
cost of the re-factoring of data compared to the eventual gains we will get?  Finally, another critical research question
is if the scientist will loose too much value from the data during the refactoring process, and whether or not we can
control this through error bounds?

The first step of re-factoring the data is to understand how much this data can be reduced.  Therefore, 
it is useful to classify scientific data into two basic categories: structured
data (in the sense that it satisfies a known or relatively simple model) or
unstructured data (in which information content follows no obvious or explicit
model).  Although scientific data generally contains random components (due to
finite precision, measurement and calibration effects etc.), scientific data is
never purely random. Broadly speaking, the path from data to knowledge consists
of extracting underlying models or patterns from the datasets, and interpreting
the resulting models.

Ideally, the scientist would perform the analysis stage {\em in situ} to
extract the relevant information and, in the unlikely event of this being
possible, would effectively circumvent the large data issue completely.  The
problem of course is that this is unlikely to be possible since, by their
nature, large scale simulations aim to discover new information that is often
hidden as higher order effects amongst the data deluge. In particular, this
means if data thinning or truncation is applied then one is in danger of
effectively throwing out the baby with the bath water. One cannot store the
entire data set in easily accessible source, due to sheer size, yet one cannot
reduce it prior to archival without risking losing the information one is
trying to extract.

@@

Our previous work~\cite{lakshminarasimhan2011compressing},~\cite{lakshminarasimhan2011compressing},~\cite{gong2012multi},
~\cite{jenkins2012byte},~\cite{gong2013parlo},~\cite{boyuka2014transparent},~\cite{tang2014improving} has focused on many
of these well known techniques to not only reduce the data but also re-organize the data. Previously we have examined:

\begin{itemize}
\item Precision based re-organization. This is where the most significant bytes of the data are
all group together from each object. This data generally will have a higher utility then the data with
the least significant bytes.  This process has the following three steps which need to be
efficiently implemented when data gets placed to the storage system and from the storage 
system. 
%
First, the data needs to be re-arranged which involves memory operations and needs
to be done in-situ. This operation involves no communication.  Data needs to be copied, so
the memory requirements are increased. At most this requires two copies an individual dataset
in memory, and we will investigate techniques to allocate and deallocate this memory if the user
will specify that the data will be overwritten after it is written to the storage system. This is often
the case for many of the data quantities written form the simulation, but there are many cases
where, for example, we want to write all of the particles from a Particle In Cell (PIC) simulation. 
Since the particles will be used later in the calculation, we need to duplicate the storage. Our
observation with working with the XGC1, GTC, Warp, and PiconGPU simulations (all leadership
class simulations) is that we can temporary increase the storage of the particles, and then 
release them since the temporary arrays used in the calculations are often freed when a PIC
iteration is finished.. 
%
\item Frequency-based re-organization. 
Another common technique to classify the importance of data is to first re-organize the data
in frequency space and then specify that as the frequency increases the importance of data
decreases.   Common technique which is used for streaming data, and data reduction techniques such as JPEG-2000~\cite{jpeg2000}
is to  code streams which have regions of interest that offer several mechanisms to support spatial random access or region of interest access at varying degrees of granularity. It is possible to store different parts of the same data using different quality.
%
This allows us to place the lowest frequency pieces in the fastest storage, and
the highest frequency either on the slowest storage tiers, or if the data sizes are prohibitively  
costly, not even write out the highest frequency. This binning of data in frequency space can
take as an input (storage bandwidth knowledge, storage size limitations, and user intentions
of how long the majority of the data will be frequently accessed.

In order to fully take advantage of frequency based re-organization we have devised a scheme where the data pre-conditioner first sorts the data
in bins of 1024 words, and then we maintain the sort order, and then we used wavelets and spline fits to then re-organize the values.
This re-organization allows us to understand that the resultant data is smoother now, and can be reduced in each one of the bins.
\end{itemize}

@@

We will measure these methods using the fusion XGC simulation and the corresponding analysis routines which they use for their largest
simulations. We will also work with Jereon Tromp's SPECFM3D code, and the fusion GTC code from Z. Lin. In all cases
we will measure their impact from their current simulations, and their projected simulations and analytics on exascale machines. Since 
all three of these codes are part of the OLCF CAAR~\cite{CAAR} project for early access, there is strong belief that the simulations will
be close to exascale simulations and we will be able to run these on prototypes of our SSIO layer to examine the full impact of data-refactoring.

\paragraph{HIgh Impact Challenges:}
Since the greatest form of data reduction on the SSIO layer would be to only write the code, and the input data,
and then recalculate all of the results, we want to understand how to write a minimal set of information, and then place an auditor 
code in a docker container so that data can be re-generated with an approximation of the calculation, but only valid over a limited amount 
of time.This new approach to tackle the data overload by
using preprocessors in which information is not simply stored in the form of
raw data, but rather in the form of appropriate parameterized algorithms. A
priori information on the nature of the data is used to choose the basic
algorithm while the actual values of the data are used to tune the parameters
used in the reconstruction procedures. We shall refer to this paradigm as an
Auditor.

>>>>

\subsection{Data Refactoring} \label{sec:data-refactor}

The classical workflow whereby the entire dataset is written to storage for
later analysis is no longer viable owing to the sheer amount of data being
generated. In future, it will be vital to take advantage of a priori
information when writing and reading from the SSIO layer in order to gain
performance and predictability, and to prioritize I/O so that the data that
is most useful for end-user can be written in the time available.

In the case of C. S. Chang described earlier, in commmon with most other
applications scientists with whom we collaborate, information is provided on
which data should be sent over to our SSIO layer in order that the best
possible information is available the subsequent analysis. The decision as how
this should be accomplished generally involves refactoring (re-organizing and
reducing the data).

In this section we describe some common methods to reduce and re-organize data.
There are many possibilities and the best choice will generally vary from case
to case but, once settled, will not change from run to run. It will be
important understand when the time and resources to required to identify and
perform the ``best methods'' outweighs the eventual gains. Another critical
research question concerns the quantification and control of the loss of
information resulting from refactoring the data and using a reduced 
dataset.

A basic issue allied to refactoring is to understand how much information is
actually present in a dataset and whether a refactoring based on a reduced
order representation might prove effective. It is useful to classify scientific
data into two basic categories: structured data (in the sense that it satisfies
a known or relatively simple model) or unstructured data (in which information
content follows no obvious or explicit model).  Although scientific data
generally contains random components (due to finite precision, measurement and
calibration effects etc.), useful scientific data is never purely random.
Broadly speaking, the path from data to knowledge consists of extracting
underlying models or patterns from the datasets, and interpreting the resulting
models.

Ideally, the scientist would perform the entire analysis stage {\em in situ} to
extract the relevant information and, in the unlikely event of this being
possible, would effectively circumvent the large data issue completely.  The
problem of course is that this is unlikely to be possible since, by their
nature, large scale simulations aim to discover new information that is often
hidden in the form of higher order effects amongst the data deluge. In
particular, this means if data thinning or truncation is applied willy nilly
then one is in danger of effectively throwing out the baby with the bath water.
One cannot store the entire data set in easily accessible source, due to sheer
size, yet one cannot reduce it prior to archival without risking losing the
information one is trying to extract.

@@

Our previous
work~\cite{lakshminarasimhan2011compressing,lakshminarasimhan2011compressing,%
gong2012multi,jenkins2012byte,gong2013parlo,boyuka2014transparent,%
tang2014improving}
has focused on many of these well known techniques to not only reduce the data
but also re-organize the data. Previously we have examined:

\begin{itemize}

\item Precision based re-organization. This is where the most significant bytes
	of the data are all group together from each object. This data
	generally will have a higher utility then the data with the least
	significant bytes.  This process has the following three steps which
	need to be efficiently implemented when data gets placed to the storage
	system and from the storage system.  First, the data needs to be
	re-arranged which involves memory operations and needs to be done
	in-situ. This operation involves no communication.  Data needs to be
	copied, so the memory requirements are increased. At most this requires
	two copies an individual dataset in memory, and we will investigate
	techniques to allocate and deallocate this memory if the user will
	specify that the data will be overwritten after it is written to the
	storage system. This is often the case for many of the data quantities
	written form the simulation, but there are many cases where, for
	example, we want to write all of the particles from a Particle In Cell
	(PIC) simulation.  Since the particles will be used later in the
	calculation, we need to duplicate the storage. Our observation with
	working with the XGC1, GTC, Warp, and PiconGPU simulations (all
	leadership class simulations) is that we can temporary increase the
	storage of the particles, and then release them since the temporary
	arrays used in the calculations are often freed when a PIC iteration is
	finished. 
%
\item Frequency-based re-organization.  Another common technique to classify
	the importance of data is to first re-organize the data in frequency
	space and then specify that as the frequency increases the importance
	of data decreases.   Common technique which is used for streaming data,
	and data reduction techniques such as JPEG-2000~\cite{jpeg2000} is to
	code streams which have regions of interest that offer several
	mechanisms to support spatial random access or region of interest
	access at varying degrees of granularity. It is possible to store
	different parts of the same data using different quality.  This allows
	us to place the lowest frequency pieces in the fastest storage, and the
	highest frequency either on the slowest storage tiers, or if the data
	sizes are prohibitively  costly, not even write out the highest
	frequency. This binning of data in frequency space can take as an input
	(storage bandwidth knowledge, storage size limitations, and user
	intentions of how long the majority of the data will be frequently
	accessed.  In order to fully take advantage of frequency based
	re-organization we have devised a scheme where the data pre-conditioner
	first sorts the data in bins of 1024 words, and then we maintain the
	sort order, and then we used wavelets and spline fits to then
	re-organize the values.  This re-organization allows us to understand
	that the resultant data is smoother now, and can be reduced in each one
	of the bins.  \end{itemize}

@@

We will measure these methods using the fusion XGC simulation and the
corresponding analysis routines which they use for their largest simulations.
We will also work with Jeroen Tromp's SPECFM3D code, and the fusion GTC code
from Z. Lin. In all cases we will measure their impact from their current
simulations, and their projected simulations and analytics on exascale
machines. Since all three of these codes are part of the OLCF CAAR~\cite{CAAR}
project for early access, there is strong belief that the simulations will be
close to exascale simulations and we will be able to run these on prototypes of
our SSIO layer to examine the full impact of data-refactoring.

\paragraph{High Impact Challenges:} Since the greatest form of data reduction
on the SSIO layer would be to only write the code, and the input data, and then
recalculate all of the results, we want to understand how to write a minimal
set of information, and then place an auditor code in a docker container so
that data can be re-generated with an approximation of the calculation, but
only valid over a limited amount of time.This new approach to tackle the data
overload by using preprocessors in which information is not simply stored in
the form of raw data, but rather in the form of appropriate parameterized
algorithms. A priori information on the nature of the data is used to choose
the basic algorithm while the actual values of the data are used to tune the
parameters used in the reconstruction procedures. We shall refer to this
paradigm as an Auditor.
